---
title: "stats480_gh2"
output: pdf_document
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

5.3
```{r}
library(tidyverse)
data <- read.delim("cars93_tab.txt")
head(data)
set.seed(100)
N <- nrow(data)
n <- 50
```

a.Select a random sample of cars from this population. Estimate the average city miles per gallon (mpg) for these cars, with a bound on the error of estimation.

Estimate of average city miles per gallon: 22.24
Confidence interval: [21.13807, 23.34193]
```{r}
sample <- sample(N, n, replace = FALSE)
sample_df <- data[sample, ]
sample_df

y <- sample_df$MPGCITY
avg_mpg <- mean(y) #estimate of avg city miles
s2 <- var(y)
varybar <- (s2/n) * (1-(n/N))
seybar <- sqrt(varybar)
error_bound_mpg <- 2 * seybar
conf_int <- avg_mpg + c(-1, 1) * error_bound_mpg
avg_mpg
conf_int
```

b. Estimate the proportion of these cars that have at least one air bag, with a bound on
the error of estimation.

Estimate of proportion of cars with at least one air bag: 0.6
Confidence interval: [0.5054268, 0.6945732]
```{r}
have_airbag <- sample_df |> filter(AIRBAGS > 0)
num_with <- nrow(have_airbag)
airbag_prop <- num_with / n #proportion with airbags = num with airbags/sample size
varphat <- ((airbag_prop * (1 - airbag_prop)) / (n-1)) * (1 - n/N)
error_bound <- 2 * sqrt(varphat)
conf_int <- airbag_prop + c(-1, 1) * error_bound
airbag_prop
conf_int
```

c. Using the data from part (a), poststratify on the car type and then estimate the aver-
age city mpg by this method.

Estimate of average after poststratification: 22.05314	
```{r}
Ni <- data |> 
  group_by(TYPE) |> 
  summarize(Ni = n())
Ni
sample_means <- sample_df |>
  group_by(TYPE) |> 
  summarise(
    mean= mean(MPGCITY),
    n_i = n()
  )
sample_means
stratified_data <- sample_means |> #combine sample_means df with Ni one
  left_join(Ni, by = "TYPE")
stratified_data

poststratified_avg_estimate <- stratified_data |>
  summarise(
    y_st = sum(Ni * mean) / N
  ) |>
  pull()
poststratified_avg_estimate
```

d. Using the data from part (b), poststratify on car type and then estimate the proportion
of cars that have at least one air bag by this method.

Estimate of proportion with at least one air bag after poststratification: 0.646496
```{r}
sample_proportions <- sample_df |>
  group_by(TYPE) |>
  summarise(
    with_airbag = sum(AIRBAGS > 0),
    n_i = n(),
    p_i = with_airbag / n_i #prop for each stratum
  )

stratified_data <- sample_proportions |>
  left_join(Ni, by = "TYPE")

poststratified_prop_estimate <- stratified_data|>
  summarise(
    p_st = sum(Ni * p_i) / N
  ) |>
  pull() #extract value from df
poststratified_prop_estimate
```

e. Comparing the above results, comment on when poststratiﬁcation might produce big
gains in terms of the error of estimation.
For both average city mpg and proportion of cars with air bags, there isn't a big difference between the estimates before and after poststratification. Poststratification might produce larger gains if the strata were more different, especially for average city mpg and proportion with air bags.  










